Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism

### 2024-07-23 index.qmd
section:preliminaries,theory; process:write

<!-- 1. What is comparative judgment? -->
In contemporary contexts, Thurstone's law of comparative judgment [-@Thurstone_1927] primarily refers to the method of *Dichotomous* Comparative Judgment [DCJ, @Pollitt_2012a; @Pollitt_2012b]. In DCJ, a judge assesses the relative manifestation of a *trait* within a pair of stimuli. This assessment results in a dichotomous value indicating which stimulus possesses a higher degree of the trait. After different judges perform multiple rounds of pairwise comparisons, an outcome vector is produced. This vector is modeled using the Bradley-Terry-Luce model [BTL, @Bradley_et_al_1952; @Luce_1959], which creates a score that corresponds with the trait of interest. This score is then used to rank the stimuli from lowest to highest or to evaluate the influence of certain variables on the stimuli's positions in the ranking.

<!-- 2. in what has been applied on?, and where is the gap? -->
DCJ has proven effective in assessing competencies and traits predominantly within the educational realm, as demonstrated by @Pollitt_2012b, @Jones_2015, @vanDaal_et_al_2016, @Bartholomew_et_al_2018, @Lesterhuis_2018, @Bartholomew_et_al_2020, and @Marshall_et_al_2020. However, its application transcends education, as exemplified by @Boonen_et_al_2020. The methodology has also evolved to include multiple, as opposed to pairwise comparisons [@Luce_1959; @Placket_1975], and to accommodate comparisons with ordinal outcomes [@Tutz_1986; @Agresti_1992]. Overall, research suggests that DCJ offers an alternative and efficient approach to measurement and evaluation, characterized by its reliability and validity [@Lesterhuis_2018; @vanDaal_2020; @Marshall_et_al_2020]. Nevertheless, despite the method's widespread use, the literature does not offer a clear representation of the plausible mechanisms that generate the DCJ data. Particularly, there is no depiction of the complexity and the underlying assumptions of the DCJ system, nor how different assessment factors can potentially influence the observed DCJ outcome.

<!-- 3. What are the factors behind DCJ? -->
According to @Verhavert_et_al_2019 and @vanDaal_2020, several assessment factors interact and influence the DCJ outcome. These factors include the number and characteristics of the stimuli, their *proximity* in terms of the assessed trait, the number of comparison per stimulus, and the pairing algorithm used. Furthermore, since the method relies on judges' assessments, the number and characteristics of judges, their *discrimination* abilities, and the number of comparisons per judge also play pivotal roles. Moreover, when the stimuli represent sub-units of higher-levels units, factors such as the number and characteristics of these units, along with their *proximity* in terms of the assessed trait, can significantly influence the outcome. For instance, @vanDaal_et_al_2016 assessed university students' skills in academic writing, utilizing multiple argumentative essays (stimuli, sub-units) originating from various students (units).

<!-- 4. Solution to the gap, final objective and relevance -->
Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the plausible mechanisms generating the DCJ outcome. This study aims to fill this gap by representing DCJ within the causal analysis framework [@Pearl_2009; @Pearl_et_al_2016]. Using Directed Acyclic Graphs [DAG, @Gross_et_al_2018; @Neal_2020], the study will construct a scientific model to elucidate the underlying assumptions of the DCJ system, providing plausible mechanisms for how the DCJ outcome could be generated. Next, using a minimal set of assumptions embedded in the framework, the study will translate the scientific model into a probabilistic statistical model. This model derives statistical estimands to draw inferences about plausible causal relationships within the DCJ system. Ultimately, this research aims to extend the law of comparative judgment initially proposed by @Thurstone_1927 and provide a sound probabilistic base for the statistical analysis of DCJ data. Consequently, this research holds significance for researchers and analysts involved in education and assessment procedures who implement or design DCJ experiments.

@@

## Causal analysis and DAGs {#sec-prelim-causal}

Causal analysis is the field of experimental design and statistics pertaining to establishing cause and effect from data.

Causal inference

Directed Acyclic Graph (DAG)

the four components of DAGs

Structural Causal Model (SCM)

the four components of DAGs in SCM notation

@@

## Where does it all fit? {#sec-theory-where}

@@

## Limitations and further research {#sec-discuss-limitations}

@@

## Why do we need to estimate judges' abilities? {#sec-appA}

>>>>

<!-- 1. What is comparative judgment? -->
In contemporary contexts, Thurstone's law of comparative judgment [-@Thurstone_1927] primarily refers to the method of *dichotomous* comparative judgment [DCJ, @Pollitt_2012a; @Pollitt_2012b]. In DCJ, a judge assesses the relative manifestation of a *trait* within a pair of stimuli. This assessment results in a dichotomous value indicating which stimulus possesses a higher degree of the trait. After different judges perform multiple rounds of pairwise comparisons, an outcome vector is produced. This vector is modeled using the Bradley-Terry-Luce model [BTL, @Bradley_et_al_1952; @Luce_1959], which creates a score that corresponds with the trait of interest. This score is then used to rank the stimuli from lowest to highest or to evaluate the influence of certain variables on the stimuli's positions in the ranking.

<!-- 2. in what has been applied on?, and where is the gap? -->
DCJ has proven effective in assessing competencies and traits predominantly within the educational realm, as demonstrated by @Pollitt_2012b, @Jones_2015, @vanDaal_et_al_2016, @Bartholomew_et_al_2018, @Lesterhuis_2018, @Bartholomew_et_al_2020, and @Marshall_et_al_2020. However, its application transcends education, as exemplified by @Boonen_et_al_2020. The methodology has also evolved to include multiple, as opposed to pairwise comparisons [@Luce_1959; @Placket_1975], and to accommodate comparisons with ordinal outcomes [@Tutz_1986; @Agresti_1992]. Overall, research suggests that DCJ offers an alternative and efficient approach to measurement and evaluation, characterized by its reliability and validity [@Lesterhuis_2018; @vanDaal_2020; @Marshall_et_al_2020]. Nevertheless, despite the method's widespread use, the literature does not offer a clear representation of the plausible mechanisms that generate DCJ data. Particularly, there is no depiction of the complexity and the underlying assumptions of the DCJ system, nor how different assessment factors can potentially influence the observed DCJ outcome.

<!-- 3. What are the factors behind DCJ? -->
According to @Verhavert_et_al_2019 and @vanDaal_2020, several assessment factors interact and influence the method's outcome. These factors include the number and characteristics of the stimuli, their *proximity* in terms of the assessed trait, the number of comparison per stimulus, and the pairing algorithm used. Furthermore, since the method relies on judges' assessments, the number and characteristics of judges, their *discrimination* abilities, and the number of comparisons per judge also play pivotal roles. Moreover, when the stimuli represent sub-units of higher-levels units, factors such as the number and characteristics of these units, along with their *proximity* in terms of the assessed trait, can significantly influence the outcome. For example, @vanDaal_et_al_2016 assessed the academic writing skills of university students (units) using multiple argumentative essays (sub-units).

<!-- 4. Solution to the gap -->
Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the mechanisms generating the DCJ outcome. This study aims to fill this gap by representing DCJ within the causal inference framework. Specifically, using the *structural approach* to causal inference [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016], the study aims to construct a scientific model. This model will elucidate the underlying assumptions of the DCJ system, providing plausible mechanisms for how the DCJ outcome is generated. Next, using a minimal set of assumptions, the study will translate the scientific model into a probabilistic statistical model. This model will produce statistical estimates to draw inferences about plausible causal relationships within the DCJ system.

<!-- 5. final objective and relevance -->
Ultimately, this research aims to extend the law of comparative judgment initially proposed by @Thurstone_1927 and provide a sound probabilistic base for the statistical analysis of DCJ data. Consequently, this research holds significance for researchers and analysts involved in education and assessment procedures who implement or design DCJ experiments.

@@

<!-- This section aims to provide the reader with the essential terms, definitions, notation, and assumptions inherent in the structural causal models (SCM) and directed acyclic graphs (DAG) framework.  -->

## The structural approach to causal inference {#sec-prelim-structural}

<!-- 1. What is causal inference and counterfactuals? -->
In statistics, *causal inference* refers to the process of identifying the causes of a phenomenon and estimating their effects using data [@Shaughnessy_et_al_2010; @Neal_2020]. Unlike classical statistical modeling, which focuses solely on summarizing data and inferring associations, causal inference provides a coherent mathematical representation for analyzing causes and counterfactuals [@Pearl_2009]. Counterfactuals reflect scenarios *contrary to fact*, where alternative *potential* outcomes have not been observed and cannot be observed [@Neal_2020; @Counterfactual_2024]. Using counterfactuals, researchers can construct a theory of the world that explains why specific causes have specific effects, and what happens in the absence of those causes. Consequently, counterfactuals represent the highest level of abstraction in causal inference [@Pearl_et_al_2018].

<!-- 2. How can we construct counterfactuals and apply causal inference? -->
Several approaches to constructing counterfactuals and applying causal inference exist, but two are particularly prominent: the potential outcomes approach and the structural approach. The potential outcomes approach, also known as the Neyman-Rubin causal model, was developed by @Neyman_et_al_1990 and @Rubin_1974. In contrast, the structural approach was pioneered by @Wright_1921, and later extended by Pearl [@Pearl_2009; @Pearl_et_al_2016]. Both approaches employ rigorous mathematical notation to characterize causal inference, but they do so in different ways [@Neal_2020]. The potential outcomes approach relies on counterfactual notation, whereas the structural approach utilizes the do-operator and Structural Causal Models [SCM, @Pearl_2009; @Pearl_et_al_2016]. Despite these differences, both approaches provide methods for using experimental and observational data to estimate causal effects [@Pearl_2010].

<!-- 3. But which approach are we going to use? -->
However, the structural approach offers a significant advantage over the potential outcomes approach by enabling the graphical representation of systems through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020]. These graphical representations provides a transparent depiction of a system's complexity, revealing its underlying assumptions and the plausible mechanisms generating the system's outcome. This capability aligns well with the aims of this study. Consequently, the structural approach will be the primary method used to address the study's research goal.

<!-- 4. But how do we move from targets of inference to estimates -->

<!-- Before tackling the main objective of this study, however, it is important to understand how researchers can move from causal estimands to statistical estimates. Causal estimands, also known as targets of inference, are the quantities that the researcher wants to estimate. Usually, these quantities are articulated through different research questions and they are depicted using the language of the Potential Outcomes framework, Neyman-Rubin causal framework [@Neyman_et_al_1990; @Rubin_1974] mathematical expectation, e.g., $E[Y_{i}(1) - Y_{i}(0)]$ indicating that a researcher is interested in estimating the *average treatment effect* (ATE) -->

<!-- This study uses the Identification-Estimation flowchart [@McElreath_2020; @Neal_2020]. The flowchart illustrates the process of moving from a target (causal) *estimand* to a corresponding *estimate*, through identification and estimation (@fig-IEflowchart).  -->

<!-- Initially, it defines the different targets of inference (i.e., the *estimands*), which form the basis for evaluating the reliability of the DCJ method. The targets  Subsequently, the research constructs a *scientific model* for the DCJ procedure, elucidating the causal assumptions and mechanisms underlying the system, aiding in drawing inferences about causal relationships from DCJ data. Next, the study translates the scientific model into a probabilistic statistical model (i.e., an *estimator*) to derive a *statistical estimand* for the various inference targets. Bayesian statistical methods offer an intuitive approach for the task. Finally, employing multiple simulated datasets for different combinations of the relevant assessment factors, the study generates multiple *estimates* or approximations for the statistical estimands. Notably, the statistical estimands in simulated data are known, enabling the study to assess the reliability of DCJ by evaluating the *proximity* of these estimands to the generated estimates, resembling a power analysis. -->

<!-- In this context, identification refers to the process of moving from a (causal) *estimand* to an equivalent *statistical estimand*, and estimation refers to the process of moving from a *statistical estimand* to an *estimate*. -->

## Graphs, DAGs and SCMs {#sec-prelim-graphs}

<!-- why to use it? -->

<!-- Causal models as a tool for causal inference -->

<!-- DAGs as a graphical representation of an SCM -->

<!-- Terms and definitions -->

<!-- the four components of DAGs -->

<!-- A *scientific (causal) model* describe the causal assumptions and mechanisms underlying a system, and facilitate inferences about causal relationships from data [@Hitchcock_2023]. -->

<!-- To aid in the translation and visualization of the assumption underlying the DCJ procedure, Directed Acyclic Graphs (DAG) and Structural Causal Models (SCM) [@Pearl_2009; @Pearl_et_al_2016; @Cinelli_et_al_2021] are employed. DAG are [(...)]{style="color:red;"}. While SCM are [(...)]{style="color:red;"}.  -->

<!-- Directed Acyclic Graph (DAG) -->

<!-- Plays an important role in effectively conveying an assumed -->
<!-- causal model and determining which variables that should be controlled for in estimation of -->
<!-- causal effects. -->

<!-- Graph means it is nodes and con- -->
<!-- nections. Directed means the connections have arrows that indicate directions of causal in- -->
<!-- fluence. And acyclic means that causes do not eventually flow back on themselves. A DAG -->
<!-- is a way of describing qualitative causal relationships among variables. -->

<!-- Structural Causal Model (SCM) -->
<!-- Causal models are mathematical models representing causal relationships within an individual system or population. They facilitate inferences about causal relationships from statistical data. They can teach us a good deal about the epistemology of causation, and about the relationship between causation and probability. They have also been applied to topics of interest to philosophers, such as the logic of counterfactuals, decision theory, and the analysis of actual causation. -->

<!-- the four components of DAGs in SCM notation -->

@@

@@

## Limitations and further research {#sec-discuss-limitations}
<!-- and most of the methods depend on unverifiable assumptions, causal inference can still identify causes and estimate their effects, assuming a set of plausibles mechanisms for a system. -->

@@

## Additional definitions {#sec-appA}

**Counterfactual:**

## Why do we need to estimate judges' abilities? {#sec-appB}
